Towards Proof-Reconstruction of Problems in Classical Propositional Logic in Agda

### 2017-09-06 paper/sections/metis-language-and-proofs.tex
[ paper ] section 2 improvements.

\textit{Input.}~The \TPTP language --which includes the first-order
form (denoted by \name{fof}) and clause normal form (denoted by \name{cnf})
formats~\cite{sutcliffe2009} -- is de
facto input standard language to encode problems for many \ATPs.
The \TPTP syntax describes a well-defined grammar to handle annotated
formulas with the following form

@@

The \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula assumes one
\name{role}, this could be an \name{axiom}, \name{conjecture},
\name{definition}, or a \name{hypothesis}. The formulas include Boolean
connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!{<=>!, $\tt\sim$) and logic
constants \verb!$true! and \verb!$false!.

For instance, we express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows.

@@

\textit{Output.}~\TSTP language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}.
A \TSTP derivation is a directed acyclic graph, a proof tree, where each leaf
is a formula from the \TPTP input. A node is a formula inferred from the parent
formulas. The root is the final derived formula. Such a derivation is a list of
annotated formulas with the form

@@

The \name{source} field is an inference record with the following pattern
\begin{verbatim}
inference(rule, useful info, parents).
\end{verbatim}

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} (see Fig.~\ref{fig:metis-inferences}) to attempt to
prove conjectures by refutation (see. e.g., the last conclusion
in the \TSTP derivation Fig.~\ref{fig:metis-example} is \emph{falsium}).

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\scriptsize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\scriptsize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\scriptsize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\scriptsize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\scriptsize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
% \caption{The \Metis' refutation tree for $p \vdash p$ from the
% derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

Despite the six rules in the \Metis' logic
kernel (see~Fig. \ref{fig:metis-inferences}), we observed in \TSTP
derivations for\CPL problems other inference rules.
They are \canonicalize, \conjunct, \negate, \simplify,
\strip and \resolve. We briefly review each of these rules
following no order except maybe by their frequency in the \TSTP derivations.

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
% \caption{Some rules of \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simplification}
\end{figure}

>>>>

\textit{Input.}~The \TPTP language --which includes the first-order
form (denoted by \name{fof}) and clause normal form (denoted by \name{cnf})
formats~\cite{sutcliffe2009} -- is de
facto input standard language to encode problems for many \ATPs.
The \TPTP syntax describes a well-defined grammar to handle annotated
formulas with the following form,

@@

The \name{language} can be \name{fof} or \name{cnf}. The \name{name}
serves to identify the formula within the problem. Each formula assumes one
\name{role}, this could be an \name{axiom}, \name{conjecture},
\name{definition}, or a \name{hypothesis}. The formulas include Boolean
connectives (\verb!&!, \verb!|!, \verb!=>!, \verb!{<=>!, \verb!~!) and logic
constants \verb!$true! and \verb!$false!.

For instance, we express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the expected
conclusion, the  \tptpconjecture \verb!~ ((p & ~ p) | (q & ~ q))!.

@@

\textit{Output.}~\TSTP language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}.
A \TSTP derivation is a directed acyclic graph, a proof tree, where each leaf
is a formula from the \TPTP input. A node is a formula inferred from the parent
formulas. The root is the final derived formula. Such a derivation is a list of
annotated formulas with the form,

@@

The \name{source} field is an inference record with the following pattern,
\begin{verbatim}
inference(rule, useful info, parents).
\end{verbatim}

@@

A proof-object delivered in a \Metis' proof encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis attempts to
prove conjectures by refutation (see. e.g., \emph{falsium} in the last conclusion
of the \TSTP derivation in Fig.~\ref{fig:metis-example}).

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\footnotesize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\footnotesize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\footnotesize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
\caption{The \Metis' refutation tree for $p \vdash p$ from the
derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

@@

We noted in \TSTP derivations that \Metis do not take into account some
inference rules for the \Metis' official documentation; we refer to
the \Metis' logic kernel. Besides the figure that shows us the six
\Metis' inference rules, its proof-objects, the \TSTP derivations,
exhibit an entirely different set of rules.

For instance, for the propositional logic fragment without equality,
we know only three rules are valid from the \Metis' Logic Kernel:
\emph{assume}, \emph{axiom}, and \emph{resolve}.
Nevertheless, as we observed in \TSTP derivations, for such a propositional
fragment, \Metis uses in his deductions steps, six rules where only the
resolve inference persists from the original list. In the case of \resolve,
it differs from its original version since, in \TSTP,  this rules comes with
an argument, a literal. The other undocumented rules found in \TSTP
derivations are: \canonicalize, \clausify, \conjunct, \negate, \simplify,
and \strip.

This lack of the \Metis' documentation was an important matter for us in the
proof reconstruction and certainty for other implementors (see a comment
related on the presenting of \name{Sledgehammer}
\info{cite here Paulson and Metis Paramodulation' paper}).
We faced the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify. For example, we believe that with these last inferences, \Metis
summarized many essential steps in the conversion or deduction from one formula
or a list of formulas to another as we show in Example X.
\unsure{Should I give an Example?}
That is inconvenient since, our proof reconstruction approach needs take into
account any assumption or consideration as ordering for literals, application
of simplification theorems, reductions, among other valid transformations.
\unsure{Should I give an Example?}
In this part, our job\improvement{job or purpose?} mainly consists of giving
a response to these rules looking at the \Metis' source code while inspecting
a representative set of solutions of \CPL problems generated by \Metis.
\unsure{Should I mention the Prop-Pack Collection?}
Therefore, this is an attempt to help and fulfill this difficulty, describing
our understanding of each rule, to later on in section
\ref{sec:proof-reconstruction}, formalize them effectively in \Agda.

@@

\[%\scalebox{0.9}{
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \bot$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \top$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \neg \varphi$}
    \UnaryInfC{$\top$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{$\varphi \vee \varphi$}
    \UnaryInfC{$\varphi$}
  \end{bprooftree}
%}
\]
\caption{Defintions of \canonicalize inference.}
\label{fig:conjunctive-disjunctive-simplification}
\end{figure}
